---
title: "IsoformSwitchAnalyzeR_practice"
output: html_document
date: "2026-05-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(IsoformSwitchAnalyzeR)
data("exampleSwitchList")
data("exampleSwitchListAnalyzed")
exampleSwitchList
```
# STEPWISE WORKFLOW
i)The raw data we get are the RNA sequencing reads in .fastq format. 
ii) Some quantification tool like Salmon maps these reads to a reference transcriptome and counts number of reads for each isoform
iii)IsoformSwitchAnalyzeR imports these counts + gene annotation

## Importing the data
```{r}
#importing salmon quantification data present in package
salmonQuant <- importIsoformExpression(
    parentDir = system.file("extdata/",package="IsoformSwitchAnalyzeR")
)

names(salmonQuant)
head(salmonQuant$abundance)
```
The salmonQuant object has two tables - counts has raw counts of reads for each isoform, abundance has normalized expression in TPM

## Building switchAnalyzeR list
```{r}
myDesign <- data.frame(
    sampleID = colnames(salmonQuant$abundance)[-1],
    condition = gsub('_.*', '', colnames(salmonQuant$abundance)[-1])
)
#take all samples except first - isoform id
#converts the conditions into plain cell names
myDesign
```
```{r}
aSwitchList <- importRdata(
    isoformCountMatrix   = salmonQuant$counts,
    isoformRepExpression = salmonQuant$abundance,
    designMatrix         = myDesign,
    isoformExonAnnoation = system.file("extdata/example.gtf.gz", package="IsoformSwitchAnalyzeR"),
    isoformNtFasta       = system.file("extdata/example_isoform_nt.fasta.gz", package="IsoformSwitchAnalyzeR"),
    showProgress = FALSE
)

aSwitchList
```
GTF file is imported from databases like Gencode, which tells where the isoform sits in the genome, what exons it has etc.
Salmon  →  tells you HOW MUCH each isoform is expressed
GTF     →  tells you WHAT each isoform looks like structurally

## Statistical testing
DEXSeq is a statistical tool to check isoform usage in bioconductor
```{r}
aSwitchList <- isoformSwitchTestDEXSeq(
    switchAnalyzeRlist = aSwitchList,
    reduceToSwitchingGenes = TRUE
)

aSwitchList
```
```{r}
head(aSwitchList$isoformFeatures[, c("gene_name", "isoform_id", "condition_1", "condition_2", "IF1", "IF2", "dIF", "isoform_switch_q_value")])
```
✅ Installed IsoformSwitchAnalyzeR
✅ Imported Salmon quantification data
✅ Built a design matrix
✅ Constructed a switchAnalyzeRlist from scratch
✅ Run statistical switch testing (DEXSeq)
✅ Understood the results biologically
IsoformSwitchPart1() can be used to do all this in one go
```{r}
### load example data
data("exampleSwitchList")

### Subset for quick runtime
exampleSwitchList <- subsetSwitchAnalyzeRlist(
    switchAnalyzeRlist = exampleSwitchList,
    subset = abs(exampleSwitchList$isoformFeatures$dIF) > 0.4
)
summary(exampleSwitchList)

```
```{r}
exampleSwitchList <- isoformSwitchAnalysisPart1(
    switchAnalyzeRlist   = exampleSwitchList,
    # pathToOutput = 'path/to/where/output/should/be/'
    outputSequences      = FALSE, # change to TRUE whan analyzing your own data 
    prepareForWebServers = FALSE  # change to TRUE if you will use webservers for external sequence analysis
)
```
Part2 is consequence analysis importing CPAT, Pfam, SignalP results
```{r}
exampleSwitchList <- isoformSwitchAnalysisPart2(
  switchAnalyzeRlist        = exampleSwitchList, 
  n                         = 10,    # if plotting was enabled, it would only output the top 10 switches
  removeNoncodinORFs        = TRUE,
  pathToCPC2resultFile      = system.file("extdata/cpc2_result.txt"         , package = "IsoformSwitchAnalyzeR"),
  pathToPFAMresultFile      = system.file("extdata/pfam_results.txt"        , package = "IsoformSwitchAnalyzeR"),
  pathToIUPred2AresultFile  = system.file("extdata/iupred2a_result.txt.gz"  , package = "IsoformSwitchAnalyzeR"),
  pathToSignalPresultFile   = system.file("extdata/signalP_results.txt"     , package = "IsoformSwitchAnalyzeR"),
  outputPlots               = FALSE  # keeps the function from outputting the plots from this example
)
extractTopSwitches(exampleSwitchList, filterForConsequences = TRUE, n=3)

```
## Visualizations
```{r}
# 1. Load the fully analyzed example
data("exampleSwitchListAnalyzed")

# 2. See the consequence summary
extractSwitchSummary(
    exampleSwitchListAnalyzed,
    filterForConsequences = TRUE
)

# 3. The famous ZAK switch plot
switchPlot(
    exampleSwitchListAnalyzed,
    gene = 'ZAK',
    condition1 = 'COAD_ctrl',
    condition2 = 'COAD_cancer',
    localTheme = theme_bw(base_size = 13)
)
```
```{r}
extractConsequenceSummary(
    exampleSwitchListAnalyzed,
    consequencesToAnalyze = 'all',
    plotGenes = FALSE,
    asFractionTotal = FALSE
)
```

